---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
knitr::opts_chunk$set(comment = "")
```

## Econ 300 Homework 1

Note: for book exercises, solutions are in appendix A. Exercises 2.25 and 2.26 are good holistic examples, but a bit difficult for this course. The "chapter exercises" for chapter 2 are similarly good but somewhat difficult questions.

## General data literacy (1.2.2)

The following book questions are examples of categorization of variables

1.3c, 1.4c, 1.5c, 1.6c, 1.10c, 1.11b 1.12b

## Quantiles and Boxplots (2.1.5)

Find the median, first quartile, third quartile, and interquartile range of the following list

```{r}
set.seed(1)
x <- c(12,3,7,1,6,5,3,8,8)
cat(x)
```


Find the median, first quartile, third quartile, and interquartile range of the following box plot

```{r,out.width="75%"}
library(ggplot2)
library(scales)
library(data.table)
set.seed(1)
y <- rnorm(100,mean=50,sd=15)
ggplot(data=data.table(y),aes(y=y)) + geom_boxplot() + theme_bw() + scale_y_continuous(breaks=pretty_breaks(10),minor_breaks=pretty_breaks(20))
```

Book Problems: 2.8

\newpage

## Histograms (2.1.3)

The following histogram has 100 total values. What percent of the data is between 30 and 40?

```{r,out.width="75%"}
hist(y,n=10)
```

What features are present in the following histogram? 

```{r, out.width="75%"}
x <- c(rgamma(100,1,1),rnorm(50,mean=5,sd=0.5))
hist(x,n=20)
```

\newpage 

For the following distribution, which is larger, the mean or the median?

```{r,out.width="75%"}
x <- c(rgamma(100,1,1))
hist(x,n=20)
```

Book Problems: 2.15, 2.16 2.14 (this one pieces together information from multiple sections)

## Mean (2.1.2)

Calculate the mean of the following data, where p is the proportion of data associated with each value

```{r}
data.table(x=c(1,2,3,4,5),p=c(.1,.2,.5,.15,.15))
```


Suppose that an outlier is added to a dataset. Which is more affected, the mean or the median?

Book Problems: 2.9

## Variance (2.1.4)

Calculate the (population) standard deviation of the following data

```{r}
cat(c(5,10,12))
```

The ACT is a standardized test with mean 18 and standard deviation 6. You score a 27 on the ACT. Calculate your z-score, i.e. your standardized test score

\newpage

Which of the following distributions has a larger standard deviation?

```{r}
x <- rnorm(1000)
y <- rnorm(1000,sd=2)

ggplot(data=data.table(x,y)) + geom_density(aes(x=x),color='red') + geom_density(aes(x=y),color='blue')+theme_bw() 
```

## Contingency Tables (2.2.1)

The following (made-up) table gives the joint distribution of econ majors (vs non-econ majors) and whether they are a dog or a cat person. 

* What percent of people prefer dogs?
* What percent of people are non-econ majors who prefer cats?
* Of people who are econ majors, what percent prefer dogs?
* Of people who prefer dogs, what percent are econ majors?

```{r}
tbl <- matrix(c(25,15,40,20,40,60,65,35,100),c(3,4))
rownames(tbl) <- c("Econ","Other","Total")
colnames(tbl) <- c("Dog","Cat","Total")
tbl
```

Book Problems: 2.22 (e is difficult)

\newpage

## Scatterplots (2.1.1)

Is the correlation between the following variables positive, negative, or approximately 0?

```{r}
x <- rnorm(100)
y <- -0.5*x + rnorm(100,sd=.5)
ggplot(data=data.table(x,y),aes(x=x,y=y)) + geom_point() + theme_bw()
```

Suppose $r=-0.9$. What percent of the variation in y is explained by x?

Book problems: 2.1, 2.2, 2.4 (positive vs negative association)

## Graphical Comparison (2.2.6)

Book problems: 2.10, 2.13